---
title: EEG Intersubject Correlation (Adapted)
description: Calculates the correlation across respondents’ EEG signals to assess the reliability of the brain response between respondents
dependencies:
  notebooks: 
    - "EEG Power Spectral Density"
  sensor: ^EEG|Raw EEG|(?i)Power\s*Bands|ActiCHamp|LSL
params:
  token: "xxxxx" 
  studyId: "xxxxx"
  stimulusId: "xxxxx"
  segmentId: "xxxxx"
  selectedSensorName: ""
  flowName: "xxxxx"
  iMotionsVersion: "xxxxx"
  window_size:
    label: "Window length [samples] for the Intersubject Correlation."
    value: 5
    input: numeric
    min: 1
  window_overlap:
    label: "Window overlap percentage [%] for the Intersubject Correlation."
    value: 50
    input: numeric
    min: 0
    max: 100 #less than 100
  quality_threshold:
    label: "Missing PSD data percentage [%] above which a respondent should be excluded."
    value: 75
    input: numeric
    min: 0
    max: 100
  multiple_device:
   label: "Consider multiple device as different respondents"
   value: FALSE
   input: checkbox
paramDetails:
  window_overlap:
    description: "Should be in the 0-100 range."
  quality_threshold:
    description: "Should be in the 0-100 range."
references:
- id: leeuwis2021
  accessed:
    - year: 2023
      month: 5
      day: 9
  author:
    - family: Leeuwis
      given: Nikki
    - family: Pistone
      given: Daniela
    - family: Flick
      given: Niels
    - family: Bommel
      given: Tom
      non-dropping-particle: van
  container-title: Frontiers in Psychology
  ISSN: 1664-1078
  issued:
    - year: 2021
  source: Frontiers
  title: 'A Sound Prediction: EEG-Based Neural Synchrony Predicts Online Music Streams'
  title-short: A Sound Prediction
  type: article-journal
  URL: https://www.frontiersin.org/articles/10.3389/fpsyg.2021.672980
  volume: '12'
output:
  html_document:
    df_print: kable
    code_folding: hide
    code_download: true
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, results = "asis")

loadPackages <- function() {
    suppressPackageStartupMessages({
        library(imotionsApi)
        library(data.table)
        library(dplyr)
        library(stringr)
        library(stringi)
        library(tidyr)
        library(zoo)
        library(kableExtra)
    })
}

loadPackages()
library(parallel)
```

```{r Parameters, purl = FALSE}
start_time <- Sys.time()

upload_data <- TRUE

# Making sure parameters are of the good format
unlockBinding("params", environment())
params$window_size <- as.numeric(params$window_size) # window length (in samples) for the ISC.
params$window_overlap <- as.numeric(params$window_overlap) # window overlap (in %) for the ISC.
params$quality_threshold <- as.numeric(params$quality_threshold) # signal quality threshold (in %) for the ISC.
```

```{r Functions common to all aggregation scripts, purl = FALSE}
#' Helper function to display information in the html report (blue rendering).
#'
#' @param text The information text to display.
#'
htmlInfo <- function(text) {
    cat("<div class='alert alert-info'><span class='glyphicon glyphicon-info-sign'></span>", paste("Info:", text),
        "</div>\n")
}


#' Helper function to display warning in the html report (orange rendering).
#'
#' @param text The warning text to display.
#'
htmlWarning <- function(text) {
    cat("<div class='alert alert-warning'><span class='glyphicon glyphicon-flash'></span>", paste("Warning:", text),
        "</div>\n")
}


#' Helper function to check if a data.table is not null and has at least one row.
#'
#' @param dt A data.table to check.
#'
#' @return A boolean set to TRUE if the data.table is not null and has at least one row.
isValid <- function(dt) {
    return(!is.null(dt) && (nrow(dt) > 0))
}


#' Helper function to find sensor matching a string pattern for further computation.
#'
#' @param study An imStudy object as returned from \code{\link{imotionsApi::imStudy}}.
#' @param respondent An imRespondent object as returned from \code{\link{imotionsApi::getRespondents}}.
#' @param stimulus An imStimulus object as returned from \code{\link{imotionsApi::getStimuli}}.
#' @param sensor_matching_pattern A string to match when retrieving a sensor.
#' @param selectedSensorName selectedSensorName from software dropdown - cloud value is empty and we use autodetection
#' @param multiple_device A boolean indicating if multiple device should be aggregated together.
#'
#' @return A data.table containing sensors that matched the string pattern.
findSensorsAggregation <- function(study, respondent, stimulus, sensor_matching_pattern, selectedSensorName,
                                   multiple_device = FALSE) {

    # get sample information for this stimulus
    sensors <- tryCatch({
        as.data.table(imotionsData::listRespondentSamples(study$connection, study$id, stimulus$id, respondent$id))
    },
    error = function(e) {
        message(paste("Could not retrieve processed data for respondent", respondent$name, "& stimulus", stimulus$name))
        NULL
    })

    if (!isValid(sensors)) return(NULL)

    sensors <- sensors[!is.na(dataUrl), ]

    # Get processed data ids (if available)
    sensor_idx <- grepl(sensor_matching_pattern, sensors$name)
    sensors <- sensors[sensor_idx, ]

    # If no sensors found matching the pattern, we won't compute any metric
    if (nrow(sensors) == 0)  return(NULL)

    if (multiple_device) {
        # in case of multiple sensors we select all data found
        sensor_idx <- seq_len(nrow(sensors))
    } else {
        # in case a device has been selected, we only run the aggregation on this one
        if (params$selectedSensorName != "") {
            # Reformat the selected device name
            device_info <- unlist(strsplit(selectedSensorName, "||", fixed = TRUE))
            instance <- unique(strsplit(paste(device_info[1], device_info[3], device_info[2]), "\\s+")[[1]])
            expected_device <- paste(instance, collapse = " ")

            # Select sensor matching the selected sensor
            sensor_idx <- grep(expected_device, str_extract(sensors$name, "(?<=\\().*(?=\\))"))
        } else {
            # if no device is selected, running it on the first processed signal found
            sensor_idx <- 1
        }
    }

    sensors <- sensors[sensor_idx, ]

    # Tag each sensorIds with the device name
    attr(sensors, "device") <- str_extract(sensors$name, "(?<=\\().*(?=\\))")
    return(sensors)
}

#' Main function to retrieve processed data for each respondent.
#'
#' @param params The list of parameters as given to the script.
#' @param study An imStudy object as returned from \code{\link{imotionsApi::imStudy}}.
#' @param respondent An imRespondent object as returned from \code{\link{imotionsApi::getRespondents}}.
#' @param stimulus An imStimulus object as returned from \code{\link{imotionsApi::getStimuli}}.
#' @param sensor_matching_pattern A string to match when retrieving a sensor.
#' @param FUN_retrieve_data Name of the function to use to retrieve data.
#'
#' @return A data.table containing the processed data for each respondent for aggregation later.
mainRetrieveData <- function(params, study, respondent, stimulus, sensor_matching_pattern, FUN_retrieve_data) {
    # Retrieve sensors available for this respondent
    sensors <- findSensorsAggregation(study, respondent, stimulus, sensor_matching_pattern, params$selectedSensorName,
                                      params$multiple_device)

    # Retrieve intervals for this stimulus
    intervals <- getRespondentIntervals(study, respondent, c("Stimulus", "Scene"))[id %in% stimulus$id, ]

    if (!isValid(sensors)) return(data.table())

    all_sensors_data <- rbindlist(by(sensors, seq_len(nrow(sensors)), function(sensor) {
            data <- FUN_retrieve_data(study, sensor, intervals)

            if (!isValid(data)) return(data.table())

            # Retrieving sensor used
            device <- attr(sensors, "device")[which(sensor$name == sensors$name)]
            data$respondentId <- paste0(respondent$id, "_", device)
            data$device <- device
            return(data)
    }))

    return(all_sensors_data)
}


#' Modified version of the bind_rows function to be able to differentiate NA values from merge induced values.
#'
#' @param ...
#'
#' @return A merged table (data.table) filled by -123455 instead of NA values.
bind_rows_fill <- function(...) {
    dots <- rlang::list2(...)

    if (length(dots) > 1) {
        headers <- unique(unlist(lapply(dots, names)))
        to_fill <- unname(which(lengths(dots) != length(headers) & lengths(dots) > 0))

        if (length(to_fill) > 0) {
            for (i in to_fill) {
                # Fill bind_rows values with -123456 to differentiate them from normal NA values and remove them later.
                columns_to_add <- setdiff(headers, names(dots[[i]]))
                suppressWarnings(dots[[i]][, eval(columns_to_add) := -123456])
            }
        }
    }

    return(bind_rows(dots))
}


#' Helper function to map and rbindlist results.
#'
#' @param data A data.table on which we want to iterate.
#' @param indices The indices to provide to the by function to iterate on.
#' @param FUN The function to apply.
#' @param keep_list A boolean indicating if single list item should be kept in a list or not.
#'
#' @return A merged table (data.table) of results.
mapBindRows <- function(data, indices, FUN, keep_list = FALSE) {
    results <- do.call(Map, c(f = bind_rows_fill, by(data, indices, function(idx) {
        res <- FUN(idx)

        if (inherits(res, "list")) {
            return(res)
        } else {
            return(list(res))
        }
    }, simplify = FALSE)))

    if (length(results) == 1 && !keep_list) {
        return(results[[1]])
    } else {
        return(results)
    }
}
```

```{r Functions specific to this export, purl = TRUE}
#' Check and modify a percentage parameter used for the ISC computation if it is out of range.
#'
#' @param percentage_param A percentage parameter as given to the script.
#' @param param_name The name of the parameter to issue a warning if needed.
#'
#' @return The modified parameter to use for the ISC computation.
checkPercentageParameter <- function(percentage_param, param_name) {
    if (!percentage_param %between% c(0, 100)) {
        percentage_param <- pmax(pmin(percentage_param, 100), 0)
        htmlWarning(paste(param_name, "is out of range, capping it to the nearest valid value:", percentage_param))
    }

    return(percentage_param)
}



#' Retrieve the PSD data for EEG Intersubject Correlation computation.
#'
#' @param study An imStudy object as returned from \code{\link{imotionsApi::imStudy}}.
#' @param sensor The sensor to retrieve.
#' @param intervals An imIntervalList object as returned from \code{\link{imotionsApi::getRespondentIntervals}}.
#'
#' @return A data.table with pre-computed PSD data.
retrievePsdDataIsc <- function(study, sensor, intervals) {
    data <- setDT(imotionsData::getSampleData(study$connection, study$id, unique(intervals$id),
                                              intervals$respondent[[1]]$id, sampleId = sensor$id))

    # Electrodes need to be re-ordered in case some data was missing for some respondents and not others
    names_electrodes <- unique(str_split(str_subset(names(data), fixed(" (dB)")), " ", simplify = TRUE)[, 1])
    names_electrodes <- str_sort(head(names_electrodes, -1), numeric = TRUE)
    ordered_names <- unlist(lapply(names_electrodes, function(x) names(data)[grepl(paste0("^", x, " "), names(data))]))

    # extract PSD part only
    data <- data[, c("TimeStamp", str_subset(ordered_names, fixed("(dB)"))), with = FALSE]
    setnames(data, 1, "Timestamp")

    # change format so the frequencies go to a separate column
    data <- pivot_longer(data, cols = starts_with(names_electrodes), names_to = c(".value", "Powerband"),
                         names_pattern = "(.*?)\\s(.*?)\\s")

    setcolorder(data, c("Timestamp", names_electrodes, "Powerband"))
    return(data)
}



#' Filter the PSD data to remove respondents with a signal quality too low.
#'
#' @param psd_data A data.table with the PSD data for all respondents that needs to be filtered.
#' @param quality_threshold A quality percentage below which respondents should be excluded from ISC.
#' @param respondents An imRespondentList object as returned from \code{\link{imotionsApi::getRespondents}}.
#'
#' @return A list containing the filtered PSD data.table containing only respondents with a good enough signal quality
#'         and a table with their quality metrics.
filterQualityPSD <- function(psd_data, quality_threshold, respondents) {
    # Filter respondents based on data quality
    frequencies <- unique(psd_data$Powerband)
    powerband_data <- psd_data[Powerband %in% frequencies[1], -c("Timestamp", "Powerband", "device")]
    ISC_respondents <- unique(powerband_data$respondentId)

    info_respondent <- stri_replace_all_fixed(paste0(str_replace_all(ISC_respondents, "_", " ("), ")"), respondents$id,
                                              respondents$name, vectorize_all = FALSE)

    # If only one device detected, we remove it from the respondent names for warning and signal quality
    device_regex <- "\\s\\([^()]+\\)"

    if (length(unique(str_extract_all(info_respondent, device_regex))) == 1) {
        info_respondent <- sub(device_regex, "\\1", info_respondent)
    }

    quality_score <- data.table(rbind(by(ISC_respondents, seq_along(ISC_respondents), function(ISC_respondent) {
        respondent_data <- powerband_data[respondentId %in% ISC_respondent, -"respondentId"]
        return(sum(is.na(respondent_data)) / prod(dim(respondent_data)) * 100)
    })))

    names(quality_score) <- info_respondent
    to_keep <- c(quality_score <= quality_threshold)

    if (any(!to_keep)) {
        if (sum(to_keep) == 0) {
            htmlWarning("All respondents had too much missing PSD data.")
        } else {
            htmlWarning(paste("Some respondents had too much missing PSD data and got excluded from the analysis:",
                        paste(info_respondent[!to_keep], collapse = ", ")))
        }
    }

    setattr(psd_data, "frequencies", frequencies)
    return(list("data" = psd_data[respondentId %in% ISC_respondents[to_keep], ], "metrics" = quality_score))
}



#' Main function to compute EEG Intersubject Correlation across respondents.
#'
#' @param params The list of parameters as given to the script.
#' @param psd_data A data.table containing the PSD processed data for each respondent.
#'
#' @return A data.table with the EEG Intersubject Correlation aggregated for each timepoint.
mainAggregationPsdIsc <- function(params, psd_data) {
    diff_ts <- median(diff(unique(psd_data$Timestamp)))
    psd_data <- psd_data[, -"Timestamp"]
    frequencies <- attr(psd_data, "frequencies")
    name_cols <- c("Timestamp", paste("Intersubject Correlation", frequencies))
    respondents <- unique(psd_data$respondentId)

    # If less than two respondents with good enough data we need to skip the ISC computation
    if (length(respondents) < 2) {
            htmlWarning(paste("Not enough respondents to compute Intersubject Correlation,",
                              "at least two respondents with valid data are needed."))

            ISC_data <- data.table(0, t(rep(NA_real_, length(frequencies))))
            setnames(ISC_data, name_cols)
            return(ISC_data)
    }

    # For the respondents with enough data quality, compute ISC
    respondents_combination <- t(combn(respondents, 2))

    # Compute sample overlap
    sample_overlap <- max(1, floor(params$window_size - params$window_overlap * params$window_size / 100))

    ISC_data <- do.call(cbind, by(frequencies, seq_along(frequencies), function(frequency) {
        psd_data <- psd_data[Powerband == frequency, -"Powerband"]

        results <- mapBindRows(respondents_combination, seq_len(nrow(respondents_combination)),
                               function(respondent_ids) {

            I <- psd_data[respondentId == respondent_ids[1], -"respondentId"]
            J <- psd_data[respondentId == respondent_ids[2], -"respondentId"]
            rows_idx <- seq_len(min(nrow(I), nrow(J)))

            if (length(rows_idx) > 0 && max(rows_idx) >= params$window_size) {
                results <- rollapply(rows_idx, width = params$window_size, by = sample_overlap, function(idx) {
                    return(abs(cor(unlist(I[idx, ]), unlist(J[idx, ]), use = "na.or.complete")))
                })
            } else {
                results <- NA_real_
            }

            results <- as.list(as.data.table(results))
            results <- lapply(results, function(x) data.table(t(x)))
            return(results)
        })

        # Set back missing windows value to NA (if any) and aggregate Intersubject Correlation
        results[results == -123456] <- NA
        return(colMeans(results, na.rm = TRUE))
    }, simplify = FALSE))

    # Generate timestamps based on the data information
    ISC_data <- as.data.table(ISC_data)
    timestamps <- seq(diff_ts * (params$window_size / 2), by = diff_ts * sample_overlap, length.out = nrow(ISC_data))

    ISC_data <- cbind(timestamps, ISC_data)
    setnames(ISC_data, name_cols)
    return(ISC_data)
}
```

```{r Retrieve data informations, purl = FALSE}
# Get information about the study ======================================================================================
connection <- imConnection(params$token)
study <- imStudy(connection, params$studyId)
stimulus <- getStimulus(study, params$stimulusId)
segment <- getSegment(study, params$segmentId)
respondents <- getRespondents(study, stimulus = stimulus, segment = segment)
```

```{r Cluster creation, purl = FALSE}
# Prepare cluster for parallel retrieving of each respondent metrics ===================================================
cl <- makeCluster(min(nrow(respondents), detectCores() - 1), type = "SOCK")
clusterExport(cl, as.list(unique(c(ls(.GlobalEnv), ls(environment())))), envir = environment())

invisible(clusterEvalQ(cl, {
    loadPackages()
    arrow::set_cpu_count(1)
}))
```

```{r Retrieve PSD data for each respondent, purl = FALSE}
# Get PSD processed data for each respondent/device ===================================================================
psd_data <- bind_rows(clusterApplyLB(cl, seq_len(nrow(respondents)), function(idx) {
    mainRetrieveData(params, study, respondents[idx, ], stimulus,
                     sensor_matching_pattern = "Power Spectral Density \\(",
                     FUN_retrieve_data = retrievePsdDataIsc)
}))

if (nrow(psd_data) == 0) {
    htmlWarning("No PSD Signal detected. Please make sure that the Power Spectral Density Notebook was
                previously run and PSD data was computed.")

    stopCluster(cl)
    knitr::knit_exit()
}
```

```{r Aggregate data and upload results, purl = FALSE}
# Verify the percentage parameters
params$window_overlap <- checkPercentageParameter(params$window_overlap, "Window overlap percentage")
params$quality_threshold <- checkPercentageParameter(params$quality_threshold, "Signal quality threshold")

if (params$multiple_device) {
    sensor_instance <- "Multiple device"
} else {
    sensor_instance <- unique(psd_data$device)
}

# Filter respondents based on data quality
quality_results <- filterQualityPSD(psd_data, params$quality_threshold, respondents)
psd_data <- quality_results$data[, -"device"]

# Aggregate PSD data to get intersubject correlation
ISC_data <- mainAggregationPsdIsc(params, psd_data)

# upload aggregated PSD data
if (upload_data) {
    group_name <- paste0(params$flowName, " (", sensor_instance, ")")
    metadata <- data.table("Group" = c("", rep(group_name, ncol(ISC_data) - 1)))

    # Add the device and respondent base to the parameters metadata so it can be retrieved by exports
    params$Device <- sensor_instance
    params$RespondentBase <- nrow(respondents)
    params$MetricsBase <- length(unique(psd_data$respondentId))
    sample_name <- script_name <- "EEG Intersubject Correlation"

    ul_data <- uploadSensorData(params, study, ISC_data, segment, sample_name, script_name, metadata, stimulus)
}
```

### Study: `r study$name`

### Stimulus: `r stimulus$name`

### Sensor: `r sensor_instance`

------------------------------------------------------------------------

### Methods

Based on the methods described in [@leeuwis2021], this R notebook has performed the following steps on the PSD data on the stimulus of interest:

-   Retrieval of the PSD data (in dB) for each respondent and filtering of respondents with a percentage of missing data above `r params$quality_threshold`%.

-   For each powerband (N_powerbands), computation of the absolute pearson correlation score for each pair of respondents (N_pairs) and each window of `r params$window_size` samples with a `r params$window_overlap`% overlap (N_windows) only taking data up to the minimal duration of exposure in case the two respondents were not exposed to the same stimulus for the same amount of time (resulting dimension N_powerbands \* N_pairs \* N_windows). Incomplete windows at the beginning and end of each stimulus are excluded from the analysis. Missing values are treated by case-wise deletion.

-   Aggregating these correlation scores across respondent pairs (N_powerbands \* N_windows).

-   Calculating the final EEG Intersubject Correlation metric for each powerband by aggregating these correlation scores over time.

------------------------------------------------------------------------

<br> <br>

### Percentage of missing PSD data (nb of samples missing / nb total of samples)

This percentage is a value ranging from 0 (good signal with no sample missing) to 100 (bad signal with all samples missing).

```{r Percentage of missing PSD data, results='asis', purl = FALSE}
if (!is.null(quality_results$metrics)) {
    percentage_valid <- quality_results$metrics
    n_tables <- ceiling(ncol(percentage_valid) / 5)

    # Splitting channels in different tables to have a nicer output
    for (i in seq(0, n_tables - 1)) {
        column_idx <- seq((i * 5) + 1, min(ncol(percentage_valid), (i * 5) + 5))

        table <- percentage_valid[, ..column_idx] %>%
            knitr::kable("html", digits = 3, align = "c") %>%
            row_spec(row = 0, background = "#EBF5FB")  %>%
            column_spec(column_idx - i * 5, width = "200px") %>%
            kable_styling(font_size = 14, position = "left", bootstrap_options = c("hover", "condensed", "responsive"),
                          full_width = FALSE)

        print(table)
    }
} else {
    htmlInfo("No raw data available.")
}

end_time <- Sys.time()
time_taken <- end_time - start_time
```

```{r End computation time, purl = FALSE}
end_time <- Sys.time()
time_taken <- end_time - start_time
```

*Computation started at `r format(start_time, usetz = TRUE)` / Notebook execution time: `r format(time_taken)`*

### References

```{r license, echo = FALSE}
# The contents of this notebook are licensed under the MIT license:
# Copyright (c) 2018-2023 iMotions
#
# Permission is hereby granted, free of charge, to any person obtaining a copy
# of this software and associated documentation files (the "Software"), to deal
# in the Software without restriction, including without limitation the rights
# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
# copies of the Software, and to permit persons to whom the Software is
# furnished to do so, subject to the following conditions:
#
# The above copyright notice and this permission notice shall be included in all
# copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
# SOFTWARE.
```
